Going Beyond Ideal Point Points: Modeling Measurement Model Measurement Error

### 2023-06-03 paper.qmd
first real draft yay

```{r}
library(targets)
library(dplyr)
library(ggplot2)
library(ggdist)
library(ggtext)
library(patchwork)
library(MetBrewer)
library(cmdstanr)
library(MCMCpack)

@@

```{=tex}
\usetikzlibrary{positioning}
\usetikzlibrary{calc}
\tikzset{
  > = stealth,
  every node/.append style = {
    draw = none
  },
  every path/.append style = {
    arrows = ->
  },
  hidden/.style = {
    draw = black,
    shape = circle,
    inner sep = 1pt
  }
}
```

```{=tex}
\begin{center}
\begin{tikzpicture}[node distance=2cm]

  % Nodes
  \node[circle, dashed, minimum size=.5cm, draw=black, fill=white] (theta) {$\theta$};
  \node[left of=theta] (etheta) {$e_{\theta}$};
  \node[below right of=theta, circle, dashed, minimum size=.5cm, draw=black, fill=white] (e) {$E$};
  \node[right of=e] (r) {$R$};
  \node[above of=r] (v) {$V$};
  \node[below of=r] (ee) {$e_E$};
  \node[above right of=theta] (thetastar) {$\theta^*$};

  % Edges
  \draw[->] (theta) -- (thetastar);
  \draw[->] (etheta) -- (thetastar);
  \draw[->] (e) -- (r);
  \draw[->] (e) -- (theta);
  \draw[->] (thetastar) -- (v);
  \draw[->] (e) -- (v);
  \draw[->] (ee) -- (r);

\end{tikzpicture}
\end{center}

```

# Introduction

@@

Theory-testing research which uses estimates from measurement models typically reduces the posterior distributions down to a single *maximum a posteriori* (MAP) value. In the case of continuous variables, researchers select some statistic of central tendency from each posterior distribution to use in subsequent analyses, such as the mean, median, or mode. This practice necessarily discards information from the full distribution. @fig-measurement-model show four hypothetical posterior distributions that may arise from a Bayesian measurement model. Despite all having the same mean of zero, higher order moments such as variance (top-right), skew (bottom-left), and kurtosis (bottom-right) can create distributions which vary to a large extent.

@@

N <- 100000
tibble(x = c(rnorm(N, sd = 1),
             rnorm(N, sd = 2),
             sn::rsn(N, alpha = -4, omega = 2, xi = 1.5),
             sn::rst(N, xi = -2.75, omega = 2, alpha = 4, nu = 1)),
       y = c(rep("A\nNormal(0, 1)", N), 
             rep("B\nNormal(0, 2)", N), 
             rep("C\nSkew Normal(1.5, 2, -4)", N), 
             rep("D\nSkew t(-2.75, 2, 4, 1)", N))) |> 
  ggplot(aes(x = x)) +
  geom_histogram(color = "white", fill = "darkcyan", alpha = .5, bins = 50) +
  geom_vline(xintercept = 0, linetype = "twodash", color = "black", linewidth = .7) +
  geom_text(aes(label = y), x = -4, y = 0) +
  facet_wrap(~ y, nrow = 2) +
  xlim(-5, 5) +
  theme_ggdist() +
  labs(y = "", x = "", caption = "Four different measurement model posterior distributions with mean zero") +
  theme(axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        text = element_text(family = "serif"))
```

In @fig-measurement-model, an estimate from distribution B should be treated as more uncertain than one from distribution A when used to test a theory. Failing to do so, as I will show, can lead to attenuation bias---or the false conclusion that the latent variable has no association with an outcome when it in fact does. In other words, the method I propose can help increase the statistical power of theory tests. Panels C and D in #fig-measurement-model show skewed distributions. Here the danger is that the skewness is caused by a third variable, which also causes the outcome of interest in the theory-testing model. This, as I will show, can lead to confounding bias if the skewness of the measurement output is not accounted for.

@@

There are two practical issues, however, with building a fully-specified joint measurement and theory-testing model. The first is computational. Bayesian statistical software uses notoriously expensive Markov Chain Monte Carlo (MCMC) sampling methods to derive its posterior distributions. Even on their own, measurement models which use MCMC can be extremely slow to sample given these types of models' high-dimensional nature. So attempting to sample from a model which also includes an arbitrarily complex theory-testing model, $f(\cdot)$, in addition to the measurement model may simply be unfeasible given the computing power that the average researcher has access to. The second challenge with the idealized joint model is that it requires researchers to write down a fully-specified measurement model, $g(\cdot)$. Compared to their theory-testing model, applied researchers likely have much less knowledge regarding the intricacies involved in estimating latent variables. Measurement models can often be challenging to fit in practice due to issues of identifiability.

My method overcomes the two problems outlined above by simplifying the measurement model step, $y^*_i \sim g(\cdot)$ in the joint model. Rather than estimating the latent variable from scratch, I take the posterior distributions already provided from previously fitted measurement models and use those as approximations in the full joint model. The measurement model $g(\cdot)$ becomes a probability distribution function with distributional parameters according to maximum likelihood estimates of the posterior. So if the posterior distribution of the measurement model appears normal, we would use $\mathbb{E}[\theta_i] \sim N(\theta_i, \sigma_i)$. The values $\mathbb{E}[\theta_i]$ and $\sigma_i$ are estimated from the measurement model's posterior distribution, which allows the true, unobserved, value of the latent variable $\theta_i$ to be estimated for each observation. If the posterior distributions from the measurement model appear skewed, or have thicker tails than a normal distribution, the distributional parameters for these distributions can be used instead. These simplifications faithfully propagate the uncertainty in the outputs of the measurement model to the theory-testing model, while also being computationally tractable and straightforward to implement.

@@

One of the most common measurement models in political science is the Bayesian Item-Response Theory (IRT) model used to measure the ideological leanings of political actors [@clinton2004; @bafumi2005]. These models assume that political ideology is a latent characteristic that lies on a single left-right dimension. Observed actions, such as voting on legislation, are used as training data ($y^*$ in @eq-ex-equation) to produce a posterior distribution of continuous values for each actor (e.g. member of Congress) on this left-right scale. [*insert footnote about how these models can measure other types of latent variables*]

Let's say we wanted to estimate the effect of legislator ideology, $\theta$, on some outcome, $y$. Using the format I developed in @eq-ex-equation, @eq-irt shows an example joint measurement and theory-testing model to answer this question using ideology estimates from an IRT model. The measurement model (bottom) predicts whether a legislator voted yes or no on a piece of legislation, $y^*_j$, using the traditional 2-parameter IRT equation $\Phi(\gamma_j\theta_i + \xi_j)$. Estimates of the parameters $\theta_i$ from this model, are then used as data in the linear regression theory-testing model (top) to estimate $\beta$---the coefficient of theoretical interest.

$$
\begin{aligned}
  y_i &\sim \text{Normal}(\alpha + \theta_i \beta, \sigma^2) \\
  y^*_{ij} &\sim \text{Bernoulli}[\Phi(\gamma_j\theta_i + \eta_i)] \\
\end{aligned}
$$ {#eq-irt}

As mentioned previously, however, estimating @eq-irt would not generally be feasible due to computational constraints. Instead, the IRT ideology measurement model should be fit beforehand. Then, for each legislator's posterior distribution of $\theta$, the values $\mathbb{E}[\theta_i]$ and $\sigma^2_{\theta i}$ should be calculated. These values can then used as data in the simplified measurement model in @eq-irt-me in order to estimate the latent $\theta_i$ for each observation.

$$
\begin{aligned}
  y_i &\sim \text{Normal}(\alpha + \theta_i \beta, \sigma^2) \\
  \mathbb{E}[\theta_i] &\sim \text{Normal}(\theta_i, \sigma^2_{\theta i})
\end{aligned}
$$ {#eq-irt-me}

@@

@@

@fig-irt-dag shows a simple DAG outlining the causal process implied by the IRT ideology measurement model. The estimated ideology variable produced by the model, $\theta^*$ is a function of an individual's true ideology, $\theta$ and measurement error, $e^{\theta}$. The important takeaway from @fig-irt-dag is that the ideology variable produced by IRT models arises causally from a combination of some latent true ideology (which we do not observe), and from measurement error (which we observe, at least partially, in the form of the posterior distribution for $\theta^*$). There is likely also considerable unobserved measurement error in these models that can be due to a variety of factors. IRT models assume each legislator's decision to vote on bill is influenced solely by their innate ideology, rather than on strategic concerns. A violation of this assumption, therefore, will produce biased estimates of $\theta^*$. There are also computational issues with fitting IRT models which may make the posterior estimates untrustworthy. For the purposes of this illustration, however, I will assume that the posterior distribution, $e^{\theta}$ for $\theta^*$ contains all relevant information about the measurement error for the true ideology $\theta$.

::: {#fig-irt-dag fig-cap="IRT Ideology Measurement Error Model"}
```{=tex}
\begin{Large}
\begin{center}
\tikz{
    \node (race_s) {$R^*$};
    \node (race_s_label) [right = .1 of race_s] {$\textcolor{teal}{\text{\normalsize Estimated Race}}$};
    \node[hidden, dashed] (race) [above = 1.5 of race_s] {$R$};
    \node (race_label) [above = .1 of race] {$\textcolor{teal}{\text{\normalsize True Race}}$};
    \node (e_race) [left = 1.5 of race_s] {$e^R$};
    \node (e_race_label) [left = 0 of e_race] {$\textcolor{teal}{\text{\normalsize Measurement Error}}$};

    \path[dashed] (race) edge (race_s);
    \path (e_race) edge (race_s);
}
\end{center}
\end{Large}
```
:::

Now consider a hypothetical theory-testing model constructed to determine whether there were racial disparities in voter turnout, $Y$ after the enactment of voter identification laws at time, $T$. Suppose we do not have data on individuals' self-reported race, and therefore must use BSIG to estimate $R^*$. @fig-race-dag-theory shows a potential causal graph for this system. The main causal effect of interest is represented by the path $R^* \longrightarrow Y$, moderated by $T$. In order to get an unbiased estimate of this causal effect we need to close all backdoor paths leading from $R^*$ to $Y$, which in this case, flows through the unobserved variable $U$. This confound represents anything that is a common cause of both the BSIG measurement error and voter turnout. @argyle2022 systematically review BSIG misclassification rates and find that socio-economic status and geography affect the amount of measurement error in the algorithm. It is also highly plausible that these variables have an independent effect on turnout rates in the population.

::: {#fig-race-dag-theory fig-cap="BSIG Measurement Error in a Hypothetical Theory-Testing Model"}
```{=tex}
\begin{Large}
\begin{center}
\tikz{
    \node (race_s) {$R^*$};
    \node[hidden, dashed] (race) [above = 1.5 of race_s] {$R$};
    \node (e_race) [left = 1.5 of race_s] {$e^R$};
    \node[hidden, dashed] (confound) [below = 1.5 of race_s] {$U$};
    \node (confound_label) [below = .1 of confound] {$\textcolor{teal}{\text{\normalsize Confound}}$};  
    \node (y) [right = 1.5 of race_s] {$Y$};
    \node (y_label) [above = .1 of y] {$\textcolor{teal}{\text{\normalsize Turnout}}$};
    \node (t) [below right = 1.5 of y] {$T$};
    \node (t_label) [below = 0 of t] {$\textcolor{teal}{\text{\normalsize Time}}$};

    \path[dashed] (race) edge (race_s);
    \path (e_race) edge (race_s);
    \path[dashed] (confound) edge [bend left=20] (e_race);
    \path[dashed] (confound) edge [bend right=20] (y);
    \path (race_s) edge (y);
    \path (t) edge (y);
}
\end{center}
\end{Large}
```
:::

For the theory-testing model in @fig-race-dag-theory it may be possible to condition on some variables in $U$ in order to obtain an unbiased estimate of $R^* \longrightarrow Y$. But with something as multi-faceted as socio-economic status, there is always the risk of residual confounding. My proposed method of building a joint measurement and theory-testing model fixes this issue by obviating the need to deal with $U$ at all. This is because the measurement error, $e^R$ in @fig-race-dag-theory is part of the backdoor path from $R^*$ to $Y$. Therefore when we explicitly incorporate $e^R$ into a model estimating $R^* \longrightarrow Y$ we can obtain an unbiased estimate of the causal effect of race on turnout.

While I have only presented a discussion of a single theory-testing model---the impact of race on turnout---incorporating $e^R$ into any model using BSIG race estimates will be valuable. Race is often thought to be an "un-caused cause" because it is assigned from an individual's birth, hence unbeholden to influence from other variables. [@sen2016]. This removes the need to worry about potential confounds because there cannot be any backdoor paths through the variable race. But as @fig-race-dag-theory shows, BSIG *estimates* of race, $R^*$ always have the potential to have open backdoor paths through measurement error $e^R$. Simply including $e^R$ directly in a joint measurement theory-testing model blocks this backdoor path going into $R^*$, which helps identify the direct effect of race on the outcome of interest.

The previous discussion of BSIG race estimates highlighted how failing to account for measurement error could lead to bias via backdoor confounding. This general issue is known as nonrandom, or unignorable, measurement error [@blalock1970]. In the political science methodology literature, methods such as multiple imputation [@blackwell2017] and sensitivity analysis [@gallop2019; @imai2010] have been developed to deal with nonrandom measurement error. My method is another way of dealing with nonrandom measurement error, but in the context where the measurement error is known and comes from the output of some measurement model.

Measurement model errors can also take the form of classical measurement error. This is where we do not assume that there is some relationship between the measurement error and the outcome of interest, rather, the errors are simply random "noise" in the measurement estimates. Classical measurement error leads to attenuation bias: a reduction of the main effect size in the theory-testing model towards zero. As I demonstrate via simulations, my method can help correct this kind of bias in theory-testing models as well.

# Simulation Study - Bayesian IRT

>>>>

```{r}
library(targets)
library(dplyr)
library(ggplot2)
library(ggdist)
library(ggtext)
library(patchwork)
library(MetBrewer)
library(cmdstanr)
library(pscl)
library(sn)
library(tidybayes)
library(MCMCpack)

@@

```{=tex}
\usetikzlibrary{positioning}
\usetikzlibrary{calc}
\tikzset{
  > = stealth,
  every node/.append style = {
    draw = none
  },
  every path/.append style = {
    arrows = ->
  },
  hidden/.style = {
    draw = black,
    shape = circle,
    inner sep = 1pt
  }
}
```
# Introduction

@@

Theory-testing research which uses estimates from measurement models typically reduces the posterior distributions down to a single value. In the case of continuous variables, researchers select some statistic of central tendency from each posterior distribution to use in subsequent analyses, such as the mean, median, or mode. This practice necessarily discards information from the full distribution. @fig-measurement-model show four hypothetical posterior distributions that may arise from a Bayesian measurement model. Despite all having the same mean of zero, higher order moments such as variance (top-right), skew (bottom-left), and kurtosis (bottom-right) can create distributions which vary to a large extent.

@@

N <- 100000
tibble(x = c(rnorm(N, sd = 1),
             rnorm(N, sd = 2),
             sn::rsn(N, alpha = -4, omega = 2, xi = 1.5),
             sn::rst(N, xi = -2.75, omega = 2, alpha = 4, nu = 1)),
       y = c(rep("A\nNormal(0, 1)", N), 
             rep("B\nNormal(0, 2)", N), 
             rep("C\nSkew Normal(1.5, 2, -4)", N), 
             rep("D\nSkew t(-2.75, 2, 4, 1)", N))) |> 
  ggplot(aes(x = x)) +
  geom_histogram(color = "white", fill = "darkcyan", alpha = .5, bins = 50) +
  geom_vline(xintercept = 0, linetype = "twodash", color = "black", linewidth = .7) +
  # geom_text(aes(label = y), x = -4, y = 0) +
  facet_wrap(~ y, nrow = 2) +
  xlim(-5, 5) +
  theme_ggdist() +
  labs(y = "", x = "", caption = "Four different measurement model posterior distributions with mean zero") +
  theme(axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        text = element_text(family = "serif"))
```

In @fig-measurement-model, an estimate from distribution B should be treated as more uncertain than one from distribution A when used to test a theory. Failing to do so, as I will show, can lead to attenuation bias---or the false conclusion that the latent variable has no association with an outcome when it in fact does. In other words, the method I propose can help increase the statistical power of theory tests. Panels C and D in @fig-measurement-model show skewed distributions. Here the danger is that the skewness is caused by a third variable, which *also* causes the outcome of interest in the theory-testing model. This, as I will show, can lead to confounding bias if the skewness of the measurement output is not accounted for.

@@

There are two practical issues, however, with building a fully-specified joint measurement and theory-testing model. The first is computational. Bayesian statistical software uses notoriously expensive Markov Chain Monte Carlo (MCMC) sampling methods to derive its posterior distributions. Even on their own, measurement models which use MCMC can be extremely slow to sample given these types of models' high-dimensional nature. So attempting to sample from a model which also includes an arbitrarily complex theory-testing model, $f(\cdot)$, in addition to the measurement model may simply be unfeasible given the computing power that the average researcher has access to. The second challenge with the idealized joint model is that it requires researchers to write down a fully-specified measurement model, $g(\cdot)$. Compared to their theory-testing model, applied researchers likely have much less knowledge regarding the intricacies involved in estimating latent variables. Because latent variables have no objective scale, measurement models can often be challenging to fit in practice due to issues of identifiability.

The method developed in this paper overcomes the two problems outlined above by simplifying the measurement model step, $y^*_i \sim g(\cdot)$ in the joint model. Rather than estimating the latent variable from scratch, I take the posterior distributions already provided from previously fitted measurement models and use those as approximations in the full joint model. The measurement model $g(\cdot)$ becomes a probability distribution function with distributional parameters according to maximum likelihood estimates of the posterior. So if the posterior distribution of the measurement model appears normal, we would use $\bar{\theta}_i \sim N(\theta_i, \sigma^2_{\theta[i]})$. The values $\bar{\theta}_i$ and $\sigma^2_{\theta[i]}$ are estimated from the measurement model's posterior distribution, which allows the true, unobserved, value of the latent variable $\theta_i$ to be estimated for each observation. If the posterior distributions from the measurement model appear skewed, or have thicker tails than a normal distribution, the distributional parameters for these distributions can be used instead. These simplifications faithfully propagate the uncertainty in the outputs of the measurement model to the theory-testing model, while also being computationally tractable and straightforward to implement.

@@

One of the most common measurement models in political science is the Bayesian Item-Response Theory (IRT) model used to measure the ideological leanings of political actors [@clinton2004; @bafumi2005]. These models assume that political ideology is a latent characteristic that lies on a single left-right dimension. Observed actions, such as voting on legislation, are used as training data ($y^*$ in @eq-ex-equation) to produce a posterior distribution of continuous values for each actor (e.g. member of Congress) on this left-right scale.

Let's say we want to estimate the effect of legislator ideology, $\theta$, on some outcome, $y$. Using the format I developed in @eq-ex-equation, @eq-irt shows an example joint measurement and theory-testing model to answer this question using ideology estimates from an IRT model. The measurement model (bottom) predicts whether a legislator voted yes or no on a piece of legislation, $y^*$, using the traditional 2-parameter IRT equation $\Phi(\gamma_j\theta_i + \xi_j)$. Estimates of the parameters $\theta_i$ from this model, are then used as data in the linear regression theory-testing model (top) to estimate $\beta_1$---the coefficient of substantive interest.

$$
\begin{aligned}
  y_i &\sim \text{Normal}(\beta_0 + \theta_i \beta_1, \sigma^2) \\
  y^*_{ij} &\sim \text{Bernoulli}[\Phi(\gamma_j\theta_i + \xi_i)] \\
\end{aligned}
$$ {#eq-irt}

As mentioned previously, however, estimating @eq-irt would not generally be feasible due to computational constraints. Instead, I propose that the IRT ideology measurement model be fit beforehand. Then, for each legislator's posterior distribution of $\theta$, the values $\bar{\theta}_i$ and $\sigma^2_{\theta[i]}$ are calculated via maximum likelihood. These values are in turn used as data in the simplified measurement model in @eq-irt-me in order to estimate the latent $\theta_i$ for each observation.

$$
\begin{aligned}
  y_i &\sim \text{Normal}(\beta_0 + \beta_1 \bar{\theta}_i, \sigma^2) \\
  \bar{\theta}_i &\sim \text{Normal}(\theta_i, \sigma^2_{\theta [i]})
\end{aligned}
$$ {#eq-irt-me}

@@

Using the logic of causal graphs, I will discuss two types of measurement model measurement error, and how the joint measurement theory-testing procedure laid out in @eq-ex-equation helps fix them. First I will consider random, or classical, measurement error. In this case the joint model will (in most cases) provide researchers with extra statistical power to test their theory. Then we will look at non-random measurement error scenarios, in which the measurement model error introduces confounding in the theory-testing model. I will show how the joint model from @eq-ex-equation ameliorates this confounding bias. For both types of measurement error, I will use simulation studies to demonstrate how effective each modeling approach is at recovering known parameter values from the theory-testing model.

If the posterior distributions for the latent parameters $\theta_i$ from the measurement model follow a normal distribution, this is a form of classical measurement error. Here we do not assume that there is some relationship between the measurement error and the outcome of interest, rather, the errors are simply random "noise" in the measurement estimates. Classical measurement error leads to attenuation bias: a reduction of the main effect size in the theory-testing model towards zero. Thus, the wider the posterior distribution is for $\bar{\theta}_i$, the more likely we are to make a False Negative error in our theory-testing model.

Let's return to the Bayesian IRT measurement model used to estimate political ideology for legislators. @fig-ideal-dag shows the causal process which produces these ideal point estimates. The observed measurements of ideology, $\bar{\theta}$ come from an unobserved latent variable plus some measurement error. At least two variables can affect the measurement error, $e^{\theta}$ in @fig-ideal-dag. First, true ideal point of the group, $\theta$ influences the amount of measurement error for estimates of the group's ideology because, as we move further from the ideological center of the scale, uncertainty increases. @fig-irt-dist shows an example of what the distribution of posterior estimates from a Bayesian IRT model look like. Groups further from the center have much wider ideal point posterior distributions. The second variable affecting $e^{\theta}$ is participation, $P$---how much a particular legislator has taken positions on bills. Legislators that signal more positions on bills will have smaller levels of measurement error compared to those that signal fewer positions because we have more data on their true ideological preferences.

::: {#fig-ideal-dag fig-cap="IRT Measurement Model"}
```{=tex}
\begin{Large}
\begin{center}
\tikz{
    \node (theta_s) {$\bar{\theta}$};
    \node (theta_label) [right = 0 of theta_s] {$\textcolor{teal}{\text{\normalsize Estimated Ideal Point}}$};
    \node[hidden, dashed] (theta) [above = 1.5 of theta_s] {$\theta$};
    \node (theta_label) [above = .1 of theta] {$\textcolor{teal}{\text{\normalsize True Ideal Point}}$};
    \node (e_theta) [left = 1.5 of theta_s] {$e^\theta$};
    \node (e_theta_label) [left = 0 of e_theta] {$\textcolor{teal}{\text{\normalsize Measurement Error}}$};
    \node (p) [below = 1.5 of e_theta] {$P$};
    \node (p_label) [below = 0 of p] {$\textcolor{teal}{\text{\normalsize Participation}}$};

There is also likely unobserved measurement error in these types of models. IRT models assume each legislator's decision to vote on bill is influenced solely by their inner ideology, rather than on strategic concerns. A violation of this assumption, therefore, will produce biased estimates of $\bar{\theta}$. There are also computational issues with fitting IRT models which can make the posterior estimates untrustworthy. For the purposes of this illustration, however, I will assume that the posterior distribution, $e^{\theta}$ for $\bar{\theta}$ contains all relevant information about the measurement error for the true ideology $\theta$.

```{r}
#| label: fig-irt-dist
#| fig-cap: "IRT Model Posterior Distributions"

set.seed(100)
N <- 300
tibble(x_meas = rnorm(N, sd = 1)) |> 
  mutate(x_sd = abs(rnorm(n(), abs(x_meas), .25))) |> 
  arrange(x_meas) |> 
  mutate(id = row_number()) |> 
  ggplot(aes(x = x_meas, y = id)) +
  geom_pointrange(aes(xmin = x_meas - x_sd, xmax = x_meas + x_sd), 
                  alpha = .35, size = .25, color = met.brewer("Isfahan1", 1)) +
  geom_pointrange(aes(xmin = x_meas - x_sd * 1.96, xmax = x_meas + x_sd * 1.96), 
                  alpha = .1, size = .25, color = met.brewer("Isfahan1", 1)) +
  # geom_point(color = "red", size = .3, alpha = .2) + 
  labs(x = expression(paste("Estimated Ideal Point, ", theta^{`*`})), y = "Groups (sorted)",
       caption = "Means and Standard Errors") +
  theme_ggdist() +
  theme(text = element_text(family = "serif"))
```

Now let's expand @fig-ideal-dag into a theory-testing model with @fig-ideal-dag-theory. While there may exist some backdoor paths through $e^{\theta}$ and $P$ in this hypothetical theory-testing model, I will assume that the outcome of interest, $Y$ is unaffected by anything other than the direct causal effect $\bar{\theta} \longrightarrow Y$.[^2] The purpose of this simplification is to highlight the consequences of random measurement error during parameter estimation of a theory-testing model.

::: {#fig-ideal-dag-theory fig-cap="IRT Measurement Model in Hypothetical Theory-Testing Model"}
```{=tex}
\begin{Large}
\begin{center}
\tikz{
    \node (theta_s) {$\bar{\theta}$};
    \node[hidden, dashed] (theta) [above = 1.5 of theta_s] {$\theta$};
    \node (e_theta) [left = 1.5 of theta_s] {$e^\theta$};
    \node (p) [below = 1.5 of e_theta] {$P$};
    \node (y) [right = 1.5 of theta_s] {$Y$};
    \node (y_label) [above = 0 of y] {$\textcolor{teal}{\text{\normalsize Outcome}}$};

### Simulation Study: Attenuation Bias

Using the generative causal model in @fig-ideal-dag-theory we can simulate data with a known parameter for the effect, $\beta_1$ of legislator ideology on the outcome $Y$. Then we fit two linear regression models to estimate this parameter. @eq-cont-no-me-model is the naive theory-testing model where $\bar{\theta}_i$ corresponds to a legislator's mean ideal point estimate from the Bayesian IRT measurement model. This is in contrast to the joint measurement theory-testing model in @eq-cont-me-model, which models $\bar{\theta}_i$ as an outcome of the true ideology $\theta_i$ (an unobserved parameter for each observation) and $\sigma^2_{\theta [i]}$ which is the observed variance of the posterior distribution for each groups' ideal point. The parameters $\theta_i$ are also given hyperpriors $\pi$ and $\tau$ for location and scale respectively. The estimates of $\theta_i$ from this simplified measurement model are then used in the linear model which predicts the outcome $y_i$. These, and all other models in this paper, are written in the probabilistic programming language Stan and fit using the No-U-Turn (NUTS) MCMC sampler [@carpenter2017].

$$
\begin{aligned}
  y_i &\sim \text{Normal}(\mu_i, \sigma^2) \\[-10pt]
  \mu_i &= \beta_0 + \beta_1 \bar{\theta}_i \\[-10pt]
  \beta_0, \beta_1 &\sim \text{Normal}(0, 2) \\[-10pt]
  \sigma &\sim \text{Half Student t}(3, 0, 2) \\
\end{aligned}
$$ {#eq-cont-no-me-model}

$$
\begin{aligned}
  y_i &\sim \text{Normal}(\mu_i, \sigma^2) \\[-10pt]
  \mu_i &= \beta_0 + \beta_1 \bar{\theta}_{i} \\[-10pt]
  \bar{\theta}_{i} &\sim \text{Normal}(\theta_i, \sigma^2_{\theta [i]}) \\[-10pt]
  \theta_i &\sim \text{Normal}(\pi, \tau) \\[-10pt]
  \beta_0, \beta_1 &\sim \text{Normal}(0, 2) \\[-10pt]
  \sigma &\sim \text{Half Student t}(3, 0, 2) \\[-10pt]
  \tau &\sim \text{Half Student t}(3, 0, 2) \\[-10pt]
  \pi &\sim \text{Normal}(0, 1)
\end{aligned}
$$ {#eq-cont-me-model}

@fig-irt-sim-comparison shows how well each model recovers the true parameter value for $\beta_1$: the effect of the true ideal point on the simulated outcome. Each model was fit 40 times across a range of increasing random error levels (shown on the horizontal axis as the correlation between the simulated true ideal point and mean measurement error value approach zero). The mean, and 89% credible interval posterior estimates of each model's $\beta_1$ parameter are plotted with a loess fit. With little-to-no measurement error (left side of the graph), both models reliably recover the true $\beta_1$ value of 1. But as the random measurement error increases, the $\beta_1$ estimates from the naive model from @eq-cont-no-me-model rapidly attenuate towards zero. This is in contrast to the estimates from the joint measurement theory-testing model in @eq-cont-me-model which remain much closer to the true $\beta_1$ value even after there is essentially zero correlation between the true ideal points and means from the ideal points with measurement error.

In addition to showing how the joint measurement theory-testing method can help avoid attenuation bias, the results from @fig-irt-sim-comparison show how this method more faithfully propagates measurement uncertainty into the theory-testing analysis. For each simulated model, the 89% credible intervals are wider for the measurement error model compared to the naive mean values model. These results demonstrate a dangerous combination of both increased bias, and increased certainty, in the theory-testing model if researchers neglect to incorporate the measurement model measurement error.

@@

The previous discussion of highlighted how failing to account for random measurement error could lead to attenuation bias. Next I will turn to the problem of measurement error-induced confounding bias. This general issue is also known as nonrandom, or unignorable, measurement error [@blalock1970]. In the political science methodology literature, methods such as multiple imputation [@blackwell2017] and sensitivity analysis [@gallop2019; @imai2010] have been developed to deal with nonrandom measurement error. My method is another way of dealing with nonrandom measurement error, but in the context where the measurement error is known and comes from the output of some measurement model.

Building off from @fig-ideal-dag-theory, let's now consider the hypothetical causal graph shown in @fig-ideal-dag-theory-c. As before, the main causal effect of interest is represented by the path $\bar{\theta} \longrightarrow Y$. In order to get an unbiased estimate of this causal effect we need to close all backdoor paths leading from $\bar{\theta}$ to $Y$, which in this case, flows through the unobserved variable $U$. This confound represents anything that is a common cause of both the IRT model measurement error and the outcome of interest.

::: {#fig-ideal-dag-theory-c fig-cap="IRT Measurement Model in Hypothetical Theory-Testing Model with Confounding"}
```{=tex}
\begin{Large}
\begin{center}
\tikz{
    \node (theta_s) {$\bar{\theta}$};
    \node[hidden, dashed] (theta) [above = 1.5 of theta_s] {$\theta$};
    \node (e_theta) [left = 1.5 of theta_s] {$e^\theta$};
    \node (p) [below = 1.5 of e_theta] {$P$};
    \node (y) [right = 1.5 of theta_s] {$Y$};
    \node[hidden, dashed] (u) [below = 1.5 of theta_s] {$U$};
    \node (u_label) [below = 0 of u] {$\textcolor{teal}{\text{\normalsize Confound}}$};

    \path (e_theta) edge (theta_s);
    \path (theta_s) edge (y);
    \path[dashed] (theta) edge (theta_s);
    \path (p) edge [bend left=0] (e_theta);
    \path[dashed] (theta) edge [bend right=30] (e_theta);
    \path[dashed] (u) edge [bend left = 30] (e_theta);
    \path[dashed] (u) edge [bend right = 30](y);
}
\end{center}
\end{Large}
```
:::

For the theory-testing model in @fig-ideal-dag-theory-c it may be possible to directly condition on some variables in $U$ in order to obtain an unbiased estimate of $\bar{\theta} \longrightarrow Y$. But with something as multi-faceted as political ideology, there is always some risk of residual confounding. My proposed method of building a joint measurement and theory-testing model fixes this issue by obviating the need to deal with $U$ at all. This is because the measurement error, $e^\theta$ in @fig-ideal-dag-theory-c is part of the backdoor path from $\bar{\theta}$ to $Y$. Therefore when we explicitly incorporate $e^\theta$ into a model estimating $\bar{\theta} \longrightarrow Y$ we can obtain an unbiased---or at least less-biased, given unobserved measurement error---estimate of the causal effect of ideology on the theoretical outcome of interest.

### Simulation Study: Confounding Bias

In order to demonstrate how the joint measurement theory-testing method I propose handles non-ignorable measurement error, I carry out a simulation study similar to that in Section 2.1.1. Given the causal graph @fig-ideal-dag-theory-c, I generate data such that $Y$ is only a function of the unobserved confound $U$. The true effect of $\theta \longrightarrow Y$ is zero in the simulation. $\bar{\theta}$ is drawn from a Skew-Normal distribution whose location parameter, $\xi$ equals the true $\theta$ value, but whose skew parameter, $\alpha$ is a function of the confound $U$. This corresponds to the $U \longrightarrow e^\theta$ path in @fig-ideal-dag-theory-c.

$$
\begin{aligned}
\frac{2}{\omega \sqrt{2 \pi}} e^{-\frac{(x-\xi)^2}{2\omega^2}} \int_{-\infty}^{\alpha\left(\frac{x-\xi}{\omega}\right)} \frac{1}{\sqrt{2 \pi}}  e^{-\frac{t^2}{2}}\ dt
\end{aligned}
$$ {#eq-skew-pdf}

@eq-skew-pdf is the probability density function for the Skew-Normal distribution. The distribution is a convolution of the Normal distribution and Half Normal (or folded Normal) distribution and has three distributional parameters: $\xi$ for location, $\omega$ for scale, and $\alpha$ for skew. When $\alpha = 0$ the distribution collapses to the Normal distribution. Unfortunately there is not a closed form solution for finding the maximum likelihood estimates of the distributional parameters so numerical methods need to be used. In practice this leads to estimation instability as $\alpha$ approaches zero [@azzalini2014]. Because of this, a choice must be made ahead of time about whether to use the Skew-Normal or regular Normal distribution for the measurement model.

After the data were generated I fit two models: the ordinary linear regression using only $\bar{\theta}$ values as used previously in the attenuation bias example (@eq-cont-no-me-model), and a modified version of @eq-cont-me-model which substitutes the Normal distribution for the Skew Normal distribution (@eq-cont-skew-model). As before, the key parameter of theoretical interest is $\beta_1$---which, according to the simulated data, should equal zero if unconfounded.

$$
\begin{aligned}
  y_i &\sim \text{Normal}(\mu_i, \sigma^2) \\[-10pt]
  \mu_i &= \beta_0 + \beta_1 \bar{\theta}_{i} \\[-10pt]
  \bar{\theta}_{i} &\sim \text{Skew Normal}(\theta_i, \omega_{\theta[i]}, \alpha_{\theta[i]}) \\[-10pt]
  \theta_i &\sim \text{Normal}(\pi, \tau) \\[-10pt]
  \beta_0, \beta_1 &\sim \text{Normal}(0, 2) \\[-10pt]
  \sigma &\sim \text{Half Student t}(3, 0, 2) \\[-10pt]
  \tau &\sim \text{Half Student t}(3, 0, 2) \\[-10pt]
  \pi &\sim \text{Normal}(0, 1)
\end{aligned}
$$ {#eq-cont-skew-model}

@fig-skew-sim-comparison shows how well each model estimates $\beta_1$. As expected, the bottom model from @eq-cont-no-me-model using only the $\bar{\theta}$ values produces a biased estimate of $\beta_1$. The open backdoor path $\bar{\theta} \longleftarrow e^\theta \longleftarrow U \longrightarrow Y$ from @fig-ideal-dag-theory-c confounds the causal effect $\theta \longrightarrow Y$ if the latent variable was measured perfectly. In contrast, the top model from @eq-cont-skew-model accurately reports a $\beta_1$ coefficient value of zero. This is because the measurement error $e^\theta$ is included in the model in the form of $\alpha_{\theta[i]}$ for each observation.[^3] As in the attenuation bias example, the naive model's posterior for $\beta_1$ is also significantly narrower compared to the measurement error model. The measurement error model is faithfully propagating the uncertainty from the measurement process into the final theory-testing analysis.

[^3]: The scale parameter $\omega_{\theta[i]}$ is also included in @eq-cont-skew-model to help avoid attenuation bias.
